A novel physical mechanism for the spontaneous formation of porous crystals via viscoelastic phase separation

### 2016-12-12 crystal-gel/nature/crystal_gel3.tex
Correct Methods section: 0.1µm is permeable to polymers "Since we have access to individual particle trajectories, we can directly assess the relative importance of these three growth channels"

The samples show significantly different behaviours in the later stages of the phase separation process.
To examine this in more detail, we plot in Fig.~\ref{fig:phase_separation}c the structure factor for the final stages of the gelation process for all state points.
Calculation of the structure factor is done with the Hanning window function, to minimize boundary effects.
The figure shows that at low wavenumber $q$, the structure factor displays fractal scaling compatible with the Guinier law, $S(q)\sim q^{-D}$.
But a difference in the fractal (or effective) dimension $D$ between the $\phi\approx 0.33$ and $c_p=0.38$ mg/g state point and other state points starts to become visible. In fact, while states
with high polymer concentration retain the exponent $D=2.5(3)$, 
which is the random gel universality class exponent, as we also observed in the early stages of the gelation process (see Fig.~\ref{fig:phase_separation}b),
the state point with low polymer concentration ($\phi\approx 0.33$ and $c_p=0.38$ mg/g) displays the biggest deviation from that exponent. In the figure we also
plot the exponent $D=3$, which corresponds to volume growth, which we will show in the following analysis to be the correct exponent for this state. 
The difference is due to the onset of crystallization in the low-polymer concentration sample. This is already evident in the high $q$ behaviour of
the structure factors (Fig.~\ref{fig:phase_separation}c), where the diffraction planes appear as sharper peaks for $\phi\approx 0.33$ and $c_p=0.38$ mg/g.

@@

A direct analysis of the colloidal positions at long times confirms that indeed the new structure obtained at
$\phi\approx 0.33$ and $c_p=0.38$ mg/g has different morphological properties than usual colloidal gels.
We directly compare the two structures in Fig.~\ref{fig:network}, where Gaussian filtering is used to depict the network structure as a continuous field.
Panel a presents the results for high
polymer concentration gel, $c_p=1.07$ mg/g; panel b presents the results for the low polymer concentration sample, $c_p=0.38$ mg/g.
The comparison shows that the new arrest mechanism produces a network structure with bigger pores, smoother surfaces, and thicker strands (panel b) compared
to the colloidal gel sample (panel a). The right column shows the individual particle positions for slabs with a thickness of 5 particle diameters. 
From the figure it is immediately clear that the network strands at low polymer concentration are crystalline. The new arrest mechanism thus
involve the formation of a crystal-gel network. In order to explain the morphological differences between colloidal gels and crystal-gels,
we now consider the kinetics of formation in our samples.

@@

\subsection{Crystal-growth channels}

While nucleation always occurs inside the liquid branches of the phase-separating network, crystal-growth can proceed through different growth channels.
In the following we will show that the growth mechanism of the crystal is not only due to filling of the spinodal liquid network, as was observed in Refs.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways} for Brownian Dynamics simulations, but also involves 
The Bergeron process and Ostwald ripening. 
In Fig.~\ref{fig:crystals}b the gyration radius of individual crystalline nuclei for the state point 
$\phi\approx 0.33$ and $c_p=0.38$ mg/g. The results show that the crystal growth follows two different scaling laws: at small crystalline sizes it scales with 
the fractal (or effective) dimension close to random percolation ($D=2.53$), while at large sizes it scales as $D=3$, as in compact crystal growth. This demonstrates that, while 
small crystalline nuclei are nucleated inside the phase-separated liquid branches, once they reach the transverse size of the liquid branch, the growth 
follows a volume growth with the formation of isotropic crystal droplets. The same scaling law was suggested in the analysis of the
structure factors, Fig.~\ref{fig:phase_separation}c, but here it is shown directly by the analysis of the gyration radius.

@@

In Fig.~\ref{fig:transitions}b, we show  the fraction of particles in each different phase for the crystallization trajectory at
$\phi\approx 0.33$ and $c_p=0.38$ mg/g, after the liquid-gas phase separation has occurred.
The process of crystallization is characterized 
by a steep decrease in liquid particles, as they transform into small crystalline nuclei inside the liquid domains. This decrease
is then accompanied by an increase in the fraction of gas particles: as the first crystals start to reach the gas phase, liquid particles evaporate 
to the gas phase due to the higher vapour pressure of the liquid phase compared to the crystalline phase. After the onset 
of a steady-state gas population, there are three growth channels for the crystal. Direct crystallization is the process by which 
crystals grow by incorporating nearby liquid particles. In the Bergeron process, liquid droplets first evaporate and the resulting gas phase 
contributes to the growth of crystalline regions. Ostwald ripening is instead the process by which small crystallites sublimate, and colloidal
particles are transferred to larger nuclei. In order to assess the relative importance of these three growth channels, we analyse the trajectory, and, for every time frame, assign each particle a state between gas, liquid and crystal. In order to minimize short-term fluctuations, the state of each colloidal particle is time averaged for $50\,\tau_{\rm B}$. We then measure the fraction of trajectories with different transition histories.
Direct crystallization accounts for 54\% of particles trajectories in which
gas or liquid particles transition to the crystal state without liquid evaporation or crystal de-sublimation. The Bergeron process accounts for 35\% of particle trajectories in which liquid particles
transition to the gas state before crystallizing. Ostwald ripening accounts for 11\% of particle trajectories in which crystal particles transition to the gas state before returning
to the crystal state. The different crystallization channels are depicted in the diagram of Fig.~\ref{fig:transitions}c.

@@

%\subsection*{Discussion and conclusions}
The first stages of gelation always involve spinodal decomposition with the formation
of liquid network by viscoelastic gas-liquid phase separation. 
% Due to hydrodynamic interaction, the network is locally stretched, and the accumulation of mechanical stress on some bonds causes some
% of them to break and the network to reorganize in a more compact and larger structure. This occurs only at low polymer concentration,
% while at high polymer concentrations the bonds are too strong to break. This will eventually dictate which arrest mechanism
% is available to the system.
Depending on the polymer concentration, there are two possible arrest mechanisms. 
(a) Crystallization, observed at $\phi\approx 0.33$ and $c_p=0.38$ mg/g, small crystalline nuclei appear inside the liquid network, reach the surface 
 of the liquid branches, and then grow by addition of particles from the gas phase. The final structure is a network of crystal droplets,
 as confirmed by the fractal dimension of the branches, the volume fraction of the particles within the branches, and bond orientational analysis. 
(b) Dynamic arrest, particle arrest when the dynamics inside the liquid branch becomes slow, which should happen at the intersection of the
 glass line with the liquid side of the coexistence curve.

@@

The process of formation of ``crystal gels'' may be generic to many other 
systems. The requirements are (i) the presence of gas-liquid phase separation below the melting point of a crystal, (ii) 
weak or little frustration against crystallization (in our case, the use of monodisperse colloids), 
(iii) dynamical slowing down in a supercooled liquid state, which is necessary to induce viscoelastic phase separation leading to the formation 
of a network structure of the minority liquid phase, and (iv) the degree of supercooling is low enough
\textcolor{blue}
{
to allow bond-breaking events that avoid the vitrification of the liquid phase.  
}
Many monoatomic and single-component molecular systems can satisfy all these conditions in a certain range of the temperature and pressure. 
This can be seen, for example, by looking at the phase diagram of a Lennard-Jones liquid \cite{lodge1997brownian}, which represents many molecular systems 
without specific directional interactions. For monoatomic systems such as noble metals, condition (iii) may not be satisfied easily. 
To access glassy slow dynamics in a liquid phase, we may need a deep quench at a high pressure. However, we note that even without strong dynamic asymmetry 
due to glassiness, bicontinuous phase separation can take place between 35-85 volume \% of the liquid phase in ordinary gas/liquid phase separation \cite{Onuki2002}; 
and, thus, porous crystalline structures can be formed, although a thin network structure may be difficult to be formed.

@@

To induce short-ranged depletion attraction, we use polystyrene (TOSOH) of molecular weight 3.8 Mdalton as non-adsorbing polymer.
Experiments are conducted at 27 $^\circ$C, some 80 $^\circ$C above the theta temperature in this solvents mixture~\cite{Royall2007}. A Flory scaling of the measurements of~\cite{lu2008gelation} yields a radius of gyration $R_g=76\pm5$ nm.

Matching phase boundaries often leads to better determination of volume fractions than colloid diameter measurements~\cite{Poon2012}. We obtain the best match between theoretical phase diagram and experimental data for $\sigma=2.25\,\mu$m and $R_g=80$ nm (see Supplementary method). Therefore the polymer-colloid size ratio is $q_R=2R_g/\sigma=0.071$ and the overlap mass fraction of polymer $2.25$~mg/g.

In the absence of salt, the Debye length is expected to reach a few $\mu$m and the (weakly) charged colloids experience a long range electrostatic repulsion. We confirm that colloids never come close enough to feel the short-ranged attraction. Screening by tetrabutylammonium bromide (Fluka) at saturated concentration brings down the Debye length to about several nanometers practically discarding the repulsion. 
Thus, salt injection (see below) can screen the Coulomb repulsion and make the polymer-induced depletion attraction effective, initiating phase separation.

Our sample cell has two layer compartments separated by membrane filters. The first layer is a sample container with thickness of 200 $\mu$m. A mixture of colloids, polymer, and solvent without salt is  set in the first layer for microscopy observation. The second layer is a salt-reservoir with a half-opened structure, which allows us to exchange or insert a reservoir solution. The volume ratio of the first and second compartments is approximately 1:100. These two compartments are separated by a membrane filter with its pore size of 0.1 $\mu$m, which passes only salt ions.

In our experiments, the reservoir solution was initially a polymer solution without salt and electrostatic repulsion by the unscreened surface charges inhibited colloidal aggregation. Under microscopy observation, we quickly exchanged the reservoir solution into a polymer solution with salt at saturated concentration (4 mM) and sealed the half-opened reservoir with cover glass to avoid evaporation of solvent. Salt diffused into the first layer within a few minutes, typically 2 minutes, screened the surface charges, and initiated colloidal aggregation. This allows us to observe the process without suffering from harmful turbulent flow.

The data are collected on a Leica SP5 confocal microscope, using 532 nm laser excitation. The scanning volume is 98 $\times$ 98 $\times$ 53 $\mu$m$^3$, which contains $\sim 10^4$ colloid particles. Our spatial resolution is 192 nm / pixel. To be able to follow individual trajectories, we perform a 3D scan every 10~s ($\approx 4\tau_{\rm B}$) at early time and every 30 s later.

>>>>

The samples show significantly different behaviours in the later stages of the phase separation process.
To examine this in more detail, we plot in Fig.~\ref{fig:phase_separation}c the structure factor for the final stages of the gelation process for all state points.
%Calculation of the structure factor is done with the Hanning window function, to minimize boundary effects.
At low wavenumber $q$, the structure factor displays fractal scaling compatible with the Guinier law, $S(q)\sim q^{-D}$.
But a difference in the fractal (or effective) dimension $D$ between the $\phi\approx 0.33$ and $c_p=0.38$ mg/g state point and other state points starts to become visible. In fact, while states
with high polymer concentration retain the exponent $D=2.5(3)$, 
which is the random gel universality class exponent, as we also observed in the early stages of the gelation process (see Fig.~\ref{fig:phase_separation}b),
the state point with low polymer concentration ($\phi\approx 0.33$ and $c_p=0.38$ mg/g) displays the biggest deviation from that exponent. In the figure we also
plot the exponent $D=3$, which corresponds to volume growth, which we will show in the following analysis to be the correct exponent for this state. 
The difference is due to the onset of crystallization in the low-polymer concentration sample. This is already evident in the high $q$ behaviour of
the structure factors (Fig.~\ref{fig:phase_separation}c), where the diffraction planes appear as sharper peaks for $\phi\approx 0.33$ and $c_p=0.38$ mg/g.

@@

A direct analysis of the colloidal positions at late times confirms that indeed the new structure obtained at
$\phi\approx 0.33$ and $c_p=0.38$ mg/g has different morphological properties than usual colloidal gels.
We directly compare the two structures in Fig.~\ref{fig:network}, where Gaussian filtering is used to depict the network structure as a continuous field.
Panel a presents the results for high
polymer concentration gel, $c_p=1.07$ mg/g; panel b presents the results for the low polymer concentration sample, $c_p=0.38$ mg/g.
The comparison shows that the new arrest mechanism produces a network structure with bigger pores, smoother surfaces, and thicker strands (panel b) compared
to the colloidal gel sample (panel a). The right column shows the individual particle positions for slabs with a thickness of 5 particle diameters. 
From the figure it is immediately clear that the network strands at low polymer concentration are crystalline. The new arrest mechanism thus
involve the formation of a crystal-gel network. In order to explain the morphological differences between colloidal gels and crystal-gels,
we now consider the kinetics of formation in our samples.

@@

\subsection{Crystal-growth routes}

While nucleation always occurs inside the liquid branches of the phase-separating network, crystal-growth can proceed through different growth \textcolor{blue}{routes}.
In the following we will show that the growth mechanism of the crystal is not only due to filling of the spinodal liquid network, as was observed in Refs.~\cite{soga1999metastable,fortini2008crystallization,perez2011pathways} for Brownian Dynamics simulations, but also involves the Bergeron process and Ostwald ripening. 
In Fig.~\ref{fig:crystals}b we show the gyration radius of individual crystalline nuclei for the state point 
$\phi\approx 0.33$ and $c_p=0.38$ mg/g. The results show that the crystal growth follows two different scaling laws: at small crystalline sizes it scales with 
the fractal (or effective) dimension close to random percolation ($D=2.53$), while at large sizes it scales as $D=3$, as in compact crystal growth. This demonstrates that, while 
small crystalline nuclei are nucleated inside the phase-separated liquid branches, once they reach the transverse size of the liquid branch, the growth 
follows a volume growth with the formation of isotropic crystal droplets. The same scaling law was suggested in the analysis of the
structure factors, Fig.~\ref{fig:phase_separation}c, but here it is shown directly by the analysis of the gyration radius.

@@

In Fig.~\ref{fig:transitions}b, we show  the fraction of particles in each different phase for the crystallization trajectory at
$\phi\approx 0.33$ and $c_p=0.38$ mg/g, after the liquid-gas phase separation has occurred.
The process of crystallization is characterized 
by a steep decrease in liquid particles, as they transform into small crystalline nuclei inside the liquid domains. This decrease
is then accompanied by an increase in the fraction of gas particles: as the first crystals start to reach the gas phase, liquid particles evaporate 
to the gas phase due to the higher vapour pressure of the liquid phase compared to the crystalline phase. 
After the onset of a steady-state gas population, there are three growth routes for the crystal. 
Direct crystallization is the process by which crystals grow by incorporating nearby liquid particles. 
In the Bergeron process, liquid droplets first evaporate and the resulting gas phase contributes to the growth of crystalline regions. Ostwald ripening is instead the process by which small crystallites sublimate, and colloidal particles are transferred to larger nuclei.
\textcolor{blue}{Since we have access to individual particle trajectories, we can directly assess the relative importance of these three growth channels.}
For every time frame, we assign each particle a state between gas, liquid and crystal. In order to minimize short-term fluctuations, the state of each colloidal particle is time averaged for $50\,\tau_{\rm B}$. We then measure the fraction of trajectories with different transition histories.
The different crystallization routes are depicted in the diagram of Fig.~\ref{fig:transitions}c.
Direct crystallization accounts for 54\% of particles trajectories in which gas or liquid particles transition to the crystal state without liquid evaporation or crystal de-sublimation. 
The Bergeron process accounts for 35\% of particle trajectories in which liquid particles transition to the gas state before crystallizing. 
Ostwald ripening accounts for 11\% of particle trajectories in which crystal particles transition to the gas state before returning to the crystal state.

@@

\subsection*{Discussion and conclusions}
The first stages of gelation always involve spinodal decomposition with the formation
of liquid network by viscoelastic gas-liquid phase separation. 
% Due to hydrodynamic interaction, the network is locally stretched, and the accumulation of mechanical stress on some bonds causes some
% of them to break and the network to reorganize in a more compact and larger structure. This occurs only at low polymer concentration,
% while at high polymer concentrations the bonds are too strong to break. This will eventually dictate which arrest mechanism
% is available to the system.
Depending on the polymer concentration, there are two possible arrest mechanisms. 
(a) Crystallization, observed at $\phi\approx 0.33$ and $c_p=0.38$ mg/g, small crystalline nuclei appear inside the liquid network, reach the surface 
 of the liquid branches, and then grow by addition of particles from the gas phase. The final structure is a network of crystal droplets,
 as confirmed by the fractal dimension of the branches, the volume fraction of the particles within the branches, and bond orientational analysis. 
(b) Dynamic arrest, particle arrest when the dynamics inside the liquid branch becomes slow, which should happen at the intersection of the
 glass line with the liquid side of the coexistence curve.

@@

The process of formation of ``crystal gels'' may be generic to many other 
systems. The requirements are (i) the presence of gas-liquid phase separation below the melting point of a crystal, (ii) 
weak or little frustration against crystallization (in our case, the use of monodisperse colloids), 
(iii) dynamical slowing down in a supercooled liquid state, which is necessary to induce viscoelastic phase separation leading to the formation 
of a network structure of the minority liquid phase, and (iv) the degree of supercooling is low enough
\textcolor{blue}{to allow bond-breaking events that avoid the vitrification of the liquid phase.}
Many monoatomic and single-component molecular systems can satisfy all these conditions in a certain range of the temperature and pressure. 
This can be seen, for example, by looking at the phase diagram of a Lennard-Jones liquid \cite{lodge1997brownian}, which represents many molecular systems 
without specific directional interactions. For monoatomic systems such as noble metals, condition (iii) may not be satisfied easily. 
To access glassy slow dynamics in a liquid phase, we may need a deep quench at a high pressure. However, we note that even without strong dynamic asymmetry 
due to glassiness, bicontinuous phase separation can take place between 35-85 volume \% of the liquid phase in ordinary gas/liquid phase separation \cite{Onuki2002}; 
and, thus, porous crystalline structures can be formed, although a thin network structure may be difficult to be formed.

@@

To induce short-ranged depletion attraction, we use polystyrene (TOSOH) as non-adsorbing polymer of molecular weight 3.8 MDa.
Experiments are conducted at 27 $^\circ$C, some 80 $^\circ$C above the theta temperature in this solvents mixture~\cite{Royall2007}. A Flory scaling of the measurements of~\cite{lu2008gelation} yields a radius of gyration $R_g=76\pm5$ nm.

Matching phase boundaries often leads to better determination of volume fractions than colloid diameter measurements~\cite{Poon2012}. We obtain the best match between theoretical phase diagram and experimental data for $\sigma=2.21\,\mu$m and $R_g=80$ nm (see Supplementary method). Therefore the polymer-colloid size ratio is $q_R=2R_g/\sigma=0.072$ and the overlap mass fraction of polymer $2.25$~mg/g.

In the absence of salt, the Debye length is expected to reach a few micrometers and the (weakly) charged colloids experience a long range electrostatic repulsion. We confirm that colloids never come close enough to feel the short-ranged attraction. Screening by tetrabutylammonium bromide (Fluka) at saturated concentration brings down the Debye length to about 100~nm practically discarding the repulsion. 
Thus, salt introduction can screen the Coulomb repulsion and make the polymer-induced depletion attraction effective, initiating phase separation.

\textcolor{blue}{To avoid solvent flow upon salt addition, our sample cell has two layer compartments separated by a membrane filter of pore size of 0.1 $\mu$m, permeable only to polymers and the salt ions. The top layer is 200 $\mu$m thick and contains the sample: a mixture of colloids, polymer, and solvent without salt. The bottom layer is a salt-reservoir with a half-opened structure, which allows us to exchange or insert a reservoir solution. The volume ratio of the first and second compartments is approximately 1:100.}

In our experiments, the reservoir solution is initially a polymer solution without salt and electrostatic repulsion by the unscreened surface charges inhibited colloidal aggregation. Under microscopy observation, we quickly exchange the reservoir solution to a polymer solution with salt at saturated concentration (4 mM) and seal the half-opened reservoir with cover glass to avoid evaporation of solvent. Salt diffuse into the first layer within a few minutes, typically 2 minutes, screen the surface charges, and initiat colloidal aggregation. This allows us to observe the process without suffering from harmful turbulent flow.

The data are collected on an upright Leica SP5 confocal microscope, using 532 nm laser excitation. The scanning volume is 98 $\times$ 98 $\times$ 53 $\mu$m$^3$, which contains $\sim 10^4$ colloid particles. Our spatial resolution is 192 nm / pixel. To be able to follow individual trajectories, we perform a 3D scan every 10~s ($\approx 4\tau_{\rm B}$) at early time and every 30 s later.
